Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-06 main.tex
Write list of players.

% Title
\title{Reinforcement Learning Produces Dominant Strategies for the
Iterated Prisoner's Dilemma}
\author{Marc Harper \and Vincent Knight} % TODO Authors?
\date{}

@@

* Update on Axelrod Library

% TODO: more

The Axelrod library \cite{knight2016open} now contains over 200 strategies,
many from the scientific literature, including classic strategies like Win Stay
Lose Shift \cite{nowak1993strategy} and previous tournament winners such as
OmegaTFT \cite{slany2007some}, Adaptive Pavlov \cite{li2007design}, and
ZDGTFT2 \cite{stewart2012extortion}.

@@

There are several previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, ashlock2015multiple, ashlock2006changes,
      ashlock2014shaped, ashlock2014evolution, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{gaudesi2016exploiting} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature.

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$. Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
First impressions matter in the IPD\@. The library includes a strategy
called EvolvedLookerUp2\_2\_2 which is among the top strategies in the library.

@@

This strategy type can be used to train arbitrary memory-$n$ strategies and a
memory one strategy was trained and is called PSO Gambler Mem 1, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSO Gamble 1 1 1.

@@

% TODO: quantify how stochastic

@@

Strategies based on artificial neural networks can also be trained with an
evolutionary algorithm. A variety of features are computed from the history
of play such as the opponents trailing moves, the total number of cooperations
of the player and the opponent, and several others, which are then input
into a feed forward neural network with one layer and user-supplied width.
An inner layer with just five nodes performs quite well in both deterministic and
noisy tournaments. The output of the ANN used in this work is deterministic and
a stochastic variant that outputs probabilities rather than exact moves could
be easily created.

% TODO: list features

@@

We used strategies based on finite state machines to create a number of
strategies. These strategies are deterministic and are efficient computational.
At each state the machine transitions to a new state and plays a specified move
depending on the opponent's last action. These strategies can encode a variety
of other strategies, including classic strategies like TitForTat, encode
handshakes, and grudging strategies that always defect after an opponent
defection.

@@

We also trained stochastic versions of finite state machine players called
hidden Markov model players or HMMs. These strategies also encode an internal
state with probabilistic transitions to other states and cooperate or defect
with various probabilities at each state. These are the best performing
stochastic strategies in the library but take longer to train due to their
stochasticity.

@@

Last but not least there are several strategies based on ensemble methods that
are common in machine learning called Meta strategies. These strategies are
composed of a team of other strategies and each is polled for its desired next
move. The ensemble then selects the next move based on some rule, such as the
consensus vote in the case of MetaMajority or the best individual performance
in the case of MetaWinner. These strategies were among the best in the library
before the inclusion of those trained by reinforcement learning.

@@

We conducted a tournament with all strategies in the Axelrod library, including
some parametrized strategies. These are listed in
Appendix~\ref{app:list_of_players}.
% TODO Add list of players
The top
11 performing strategies by median payoff are all strategies trained to maximize
% TODO Potentially modify this as more repetitions are analysed.
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is
Desired Belief Strategy (DBS)
% TODO Add reference for DBS
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection and Omega Tit For Tat.
% TODO Add a reference for Omega TfT
All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:

@@

% Table of best strategies
\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments)}
        \label{tbl:standard_score}
\end{table}

@@

Pairwise payoff results are given as a heatmap (Figure~\ref{fig:standard_heatmap})
which shows that many strategies achieve mutual cooperation with each other and
that the top performing strategies never defect first but are able to exploit
weaker strategies that attempt to defect.

@@

Finally, Table~\ref{tbl:standard_ranks} and
Figure~\ref{fig:standard_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. Whilst there is some
stochasticity, the top three strategies almost always rank in the top three. For
example, the worst that the Evolved Lookerup 2 2 2 ranks in a given tournament
is 7th.

@@

% Table of best strategies
\begin{table}[!hbtp]
    \centering
        \input{./assets/noisy_top_15_scores_summary.tex}
        \caption{Noisy (5\%) Tournament: Mean score per turn of top 15 strategies
        (ranked by median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)}
        \label{tbl:noisy_score}
\end{table}

@@

As shown in Table~\ref{tbl:noisy_wins}, the top ranking strategies win a larger
number of matches in the presence of noise. For example Spiteful Tit For Tat in
one tournament won almost all it's matches (167).

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see, that the stochasticity
of the ranks understandably increases the DBS strategy never ranks lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

Using the method of fingerprinting described in \cite{ashlock2008fingerprinting}
\cite{ashlock2009fingerprint}, we can compare
strategies. For the top performing noisy strategies there is a striking similarity
in the fingerprints.

@@

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler particle swarm algorithms. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approximately
200 generations or until strategies no longer improve significantly.

% Todo: Diagram with ANN training performance?

All training code is available on github. There are objective functions for
* total payoff
* total payoff difference
* total Moran process wins (fixation probability)

New strategies can be easily trained with variations including noise, spatial
structure, and probabilistically ending matches.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher median
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategy did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

@@

Of the strategies trained to maximize their median score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process for maximizing
payout does not
lead to exploitative zero determinant strategies, which may be a result of the
collection of training strategies, many of which retaliate harshly.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0) that did not include DBS
and several other opponents. The precise parameters that are optimal will depend
on the pool of opponents. Moreover we have not extensively trained strategies to
determine the minimum parameters that are sufficient -- neural networks with
fewer nodes and features and finite state machines with fewer states may suffice.
See \cite{ashlock2013impact} for discussion of resource availability for IPD strategies.

@@

% TODO: Full strategy list

@@

>>>>

% Title
\title{Reinforcement Learning Produces Dominant Strategies for the
Iterated Prisoner's Dilemma}
\author{Marc Harper \and Vincent Knight} % TODO Authors
\date{}

@@

The Axelrod library \cite{knight2016open} is an open source software for
conducting iterated prisoner's dilemma (IPD) research with reproducibilty as a
principal goal. Written in the Python programming language, to date over the
library contains source code contibuted by over 50 individuals from a variety
of geographic locations and technical backgrounds. The library is supported by
a comprehensive test suite that covers all the intended behaviors of the
strategies in the library, as well as the features that conduct matches,
tournaments, and population dynamics.

As of version 3.0.0, the library contains over 200 strategies,
many from the scientific literature, including classic strategies like Win Stay
Lose Shift \cite{nowak1993strategy} and previous tournament winners such as
OmegaTFT \cite{slany2007some}, Adaptive Pavlov \cite{li2007design}, and
ZDGTFT2 \cite{stewart2012extortion}.

@@

In this work we utilize the collection of strategies in the Axelrod library
to train new strategies specifically to win IPD tournaments. We train these
strategies using generic strategy archetypes based on e.g. finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, ashlock2015multiple, ashlock2006changes,
      ashlock2014shaped, ashlock2014evolution, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{gaudesi2016exploiting} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature. Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented and documented. Moreover, as of the time of writing,
we claim that this work contains the best known strategies for the iterated
prisoner's dilemma.

@@

The first strategy trained with reinforcement learning in the library is based
on lookup tables. The strategy encodes a set of deterministic responses
based on the opponent's first $n_1$ moves, the opponent's last $m_1$ moves, and
the players last $m_2$ moves. If $n_1 > 0$ then the player has infinite memory
depth, otherwise it has depth $\max{m_1, m_2}$. Although we tried various
combinations of $n_1, m_1,$ and $m_2$, the best performance at the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
The library includes a strategy
called EvolvedLookerUp2\_2\_2 which is among the top strategies in the library.

@@

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSO Gambler Mem 1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSO Gambler 1 1 1.

@@

@@

Strategies based on artificial neural networks can also be trained with an
evolutionary algorithm. A variety of features are computed from the history
of play such as the opponents trailing moves, the total number of cooperations
of the player and the opponent, and several others, which are then input
into a feed forward neural network with one layer and user-supplied width.
An inner layer with just five nodes performs quite well in both deterministic and
noisy tournaments. The output of the ANN used in this work is deterministic;
a stochastic variant that outputs probabilities rather than exact moves could
be easily created.

@@

We used strategies based on finite state machines to create a number of
strategies. These strategies are deterministic and are efficient computationally.
In each round of play the strategy selects an action based on the current state
and the opponent's last action, transitioning to a new state for the next round.
These strategies can encode a variety
of other strategies, including classic strategies like TitForTat, 
handshake strategies, and grudging strategies that always defect after
an opponent defection.

@@

We also trained stochastic versions of finite state machine players called
hidden Markov model players or HMMs. Like the finite state machines, these
strategies also encode an internal
state with probabilistic transitions based on the prior round of play to other
states and cooperate or defect with various probabilities at each state. These
are the best performing stochastic strategies in the library but take longer
to train due to their stochasticity.

@@

Last but not least there are several strategies based on ensemble methods that
are common in machine learning called Meta strategies. These strategies are
composed of a team of other strategies. In each round, each member of the team
is polled for its desired next
move. The ensemble then selects the next move based on a rule, such as the
consensus vote in the case of MetaMajority or the best individual performance
in the case of MetaWinner. These strategies were among the best in the library
before the inclusion of those trained by reinforcement learning.

@@

We conducted a tournament with a large collection of strategies from the Axelrod
library, including some additional parametrized strategies (e.g. various parameter
choices for Generous Tit For Tat). These are listed in
Appendix~\ref{app:list_of_players}.
The top 11 performing strategies by median payoff are all strategies trained to maximize
% TODO Potentially modify this as more repetitions are analysed.
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS)
% TODO Add reference for DBS
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{Slany2007}.
All strategies in the tournament follow a simple set of
rules in accordance with earlier tournaments:

@@

Any strategy that does not follow these rules, such as a strategy that defects
on the last round of play, was omitted from the tournament presented here (but
not from the training pool).

% Table of best strategies
\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the machine learning strategies. ~$^{*}$
        indicates that the strategy was trained.}
        \label{tbl:standard_score}
\end{table}

@@

Pairwise payoff results are given as a heatmap (Figure~\ref{fig:standard_heatmap})
which shows that many strategies achieve mutual cooperation. The top performing
strategies never defect first yet are able to exploit weaker strategies that
attempt to defect.

@@

Finally, Table~\ref{tbl:standard_ranks} and
Figure~\ref{fig:standard_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. Whilst there is some
stochasticity, the top three strategies almost always rank in the top three. For
example, the worst that the Evolved Lookerup 2 2 2 ranks in a given tournament
is 8th.

@@

% Table of best strategies
\begin{table}[!hbtp]
    \centering
        \input{./assets/noisy_top_15_scores_summary.tex}
        \caption{Noisy (5\%) Tournament: Mean score per turn of top 15 strategies
        (ranked by median over
        \protect\input{./assets/noisy_number_of_repetitions.tex}tournaments)
        ~$^{*}$ indicates that the strategy was trained.}
        \label{tbl:noisy_score}
\end{table}

@@

As shown in Table~\ref{tbl:noisy_wins}, the top ranking strategies win a larger
number of matches in the presence of noise. For example Spiteful Tit For Tat in
one tournament won almost all its matches (167).

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases the DBS strategy never ranks lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

Again we compare the fingerprints. For the top performing noisy strategies there
is a striking similarity in the fingerprints which indicates that the strategies
may behave similarly in principle.

@@

We trained a variety of strategies using evolutionary algorithms, and in the
case of PSO Gambler using a particle swarm algorithm. The evolutionary algorithms
used standard techniques, varying strategies by mutation and crossover, and
evaluating the performance against each opponent for many repetitions. The
best performing strategies in each generation are persisted, variants created,
and objective functions computed again. This process continues for approximately
200 generations or until strategies no longer improve significantly.

% TODO: Diagram with ANN training performance?

All training code is available on github. There are objective functions for
\begin{itemize}
 \item total or mean payoff
 \item total or mean payoff difference (unused in this work)
 \item total Moran process wins (fixation probability)
\end{itemize}

These objectives can be easily modified to suit other purposes. New strategies
can be easily trained with variations including noise, spatial structure, and
probabilistically ending matches.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. A simple evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategies did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well. Single layer neural networks also performed well
though these had some aspect of human involvement in the selection of features.
The success of the other strategy types suggests that a deeper network that
incorporates feature engineering would likely also perform well.

@@

Of the strategies trained to maximize their median score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable evolutionary equilibrium for the prisoner's
dilemma. Specifically we note that the reinforcement learning process for maximizing
payout does not lead to exploitative zero determinant strategies, which may also
be a result of the collection of training strategies, of which several retaliate harshly.

We take the liberty of generalizing from the results of this study. For the trained
strategies utilizing look up tables we generally found those that incorporate
one of more of the initial rounds of play outperformed those that did not. The
strategies based on neural networks and finite state machines also are able to
condition throughout a match on the first rounds of play. Accordingly, we conclude
that first impressions matter in the IPD\@. The best strategies are nice (never
defecting first) and this property could be further investigated with the library
in future work by e.g. forcing all strategies to defect on the first round.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0) that did not include DBS
and several other opponents. The precise parameters that are optimal will depend
on the pool of opponents. Moreover we have not extensively trained strategies to
determine the minimum parameters that are sufficient -- neural networks with
fewer nodes and features and finite state machines with fewer states may suffice.
See \cite{ashlock2013impact} for discussion of resource availability for IPD
strategies. It may be possible to train strategies more effective in noisy tournaments
than DBS.

@@

@@

\appendix

\section{List of players}\label{app:list_of_players}

\begin{multicols}{2}
	\begin{enumerate}
		\input{./assets/list_of_players.tex}
	\end{enumerate}
\end{multicols}
